Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-01-23 draft.tex
Update draft after another discussion with Fred

\begin{abstract}
  \fred{Our goal is to obtain a complete set} of angular observables
  arising in a generic multi-body process without the need to carry
  out a likelihood fit of the angular distribution to the measured
  events. Instead, we \fred{apply the method of moments} that relies
  both on the orthogonality of angular functions and the estimation of
  integrals by Monte Carlo techniques. \fred{The main advantage of
    this method is that the joint distribution of all observables
    given the data can be easily extracted, even for very few
    events. Our main contribution is to show explicitly how systematic
    uncertainties from detector-resolution and acceptance effects can
    be incorporated in a general algorithm. Finally, we present the
    necessary process-dependent formulae needed for direct
    implementation of the method for several rare decays of
    interest.}
\end{abstract}

@@

\begin{figure}
    \includegraphics[width=.6\textwidth]{fig-topology.pdf}
    \caption{Decay topology for decays $B\to P_1 P_2 \ell_1 \bar\ell_2$. \label{fig:topology}\fred{We should use the same symbol $\theta$ as in text}}
\end{figure}

@@

\subsection{Recipe for \fred{Including} Detector Effects}
\label{sec:systematics:acceptance}

@@

The accuracy of the unfolding process as outlined above critically depends on both the accuracy
of the detector simulation, as well as the uncertainties induced by the MC estimates. While
considerations of any detector simulation is beyond the scope of this work, we can, however,
comment on the MC-induced uncertainties. As usual, one needs to find a balance between
compute time and accuracy. For a uniangular distribution and a total of $6\cdot 10^6$ MC
samples (i.e., $2 \cdot 10^6$ MC samples per row of $M_{ij}$), we find that the error on the mean is
smaller than $6\cdot 10^{-4}$. Let us conclude this section by commenting that parts of the unfolding
matrix are universal in a sense. Specifically, they can be reused in analyses with a similar underlying decay.
Therefore, computing resources spent on improving its accuracy are not wasted.

@@

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S5.png}
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S7.png}
            \caption{Uncertainty of the angular observables $S_5$ and $S_7$, extracted from simulated events of the decay $B\to K^*\ell^+\ell^-$ and as a function of the number of simulated events $K$. The red curve represents a fit to the function given in \refeq{unc-on-mean}.}
            \label{fig:errors}
            \fred{Plots convey exactly was is said in words. I see they are useful for a talk, but do we really need them in the write-up?}
\end{figure}

@@

From the positivity of amplitudes and unitarity, one immediately finds
\begin{equation}
    \int \rmdx{\cos\theta}  \, |A_i(\theta)|^2 = \sum_{l_i = 0}^\infty |A_i^{(l_i)}|^2 n_{l_i, m_i} < \infty\,,
\end{equation}
where we introduce $n_{l,m}$ via the scalar product of two associated Legendre polynomials,
\begin{equation}
    \label{eq:legendre-scalar-product}
    n_{l, m} \delta_{l, l'} \equiv \int_{-1}^1 \rmdx{\cos\theta} p_{l}^{(m)}(\cos\theta) p_{l'}^{(m)} (\cos\theta)  = \frac{2}{(2 l + 1)} \frac{(l + m)!}{(l - m)!} \delta_{l, l'}\,.
\end{equation}
Using the large-$l$ behaviour of $n_{l,m}$ one finds that with
increasing $l_i$ the partial-wave amplitudes must asymptotically fall
off as
\begin{equation}
    |A_i^{(l_i)}| \Big|_{l_i \gg m_i} \sim \frac{1}{\sqrt{l_i}}\,.
\end{equation}
This motivates to cut off the partial-wave expansion at some arbitrary angular momentum $L$.\\

@@

In the last step, we use Gaunt's formula \cite{Gaunt:1929} to integrate a triple product of associated Legendre polynomials,
\begin{equation}
\begin{aligned}
    T_{l_1,l_2,j}^{(m_1,m_2)}
        & = \int_{-1}^{+1} \rmdx{\cos\theta} p_{j}^{(m_1 + m_2)}(\cos\theta) p_{l_1}^{(m_1)}(\cos\theta) p_{l_2}^{(m_2)}(\cos\theta)\\
        & = (-1)^{s - l_1 - m_2} \frac{2 (l_1 + m_1)! (l_2 + m_2)! (2s - 2 l_2)! s!}{(l_1 - m_1)! (s - j)! (s - l_1)! (s - l_2)! (2s + 1)!}\\
        & \times \sum_{t=p}^q (-1)^t \frac{(j + m + t)!(l_1 + l_2 - m - t)!}{t! (j - m - t)! (l_1 - l_2 + m + t)! (l_2 - m_2 - t)!}\,,
\end{aligned}
\end{equation}
where
\begin{equation}
\begin{aligned}
    m & = m_1 + m_2\,, &
    m_1 & \geq m_2\,,  \\
    j, l_1, l_2 & \geq 0\,, &
    m, m_1, m_2 & \geq 0\,,
\end{aligned}
\end{equation}
and
\begin{equation}
\begin{aligned}
    s & = \frac{j + l_1 + l_2}{2}\,, &
    p & = \max(0, l_2 - l_1 - m)\,, &
    q & = \min(l_1 + l_2 - m, j - m, l_2 - m_2)\,.
\end{aligned}
\end{equation}
The necessary conditions for $T \neq 0$ are
\begin{equation}
    \label{eq:angular-momentum-addition}
    s \in \mathbb{N}\qquad \wedge \qquad l_1 - l_2 \leq l \leq l_1 + l_2,.
\end{equation}
The latter condition is well known from the addition rules of angular momenta. Note, however, that
the sum in \refeq{def-partial-wave-observable} goes to infinitely high angular momenta $l_1$ and $l_2$. As a consequence
of this and of \refeq{angular-momentum-addition}, the angular observables $P_{j,m}$
consist of sums with infinitely many terms. It is then up to theoretical analyses to
estimate or calculate the impact of individual contributions, with or without the cutoff $L$.
We close by remarking that for any angular-momentum cutoff $l < L$\fred{Do you mean $L < \infty$?}, one can find an upper limit for the squared sum of all remaining partial waves through a comparison with the inclusive decay
width.

>>>>

\begin{abstract}
  Our goal is to obtain a complete set of angular observables
  arising in a generic multi-body process without the need to carry
  out a likelihood fit of the angular distribution to the measured
  events. Instead, we apply the method of moments that relies
  both on the orthogonality of angular functions and the estimation of
  integrals by Monte Carlo techniques. The main advantage of
  this method is that the joint distribution of all observables
  given the data can be easily extracted, even for very few
  events. Our main contribution is to show explicitly how systematic
  uncertainties from detector-resolution and acceptance effects can
  be incorporated in a general algorithm. Finally, we present the
  necessary process-dependent formulae needed for direct
  implementation of the method for several rare decays of
  interest.
\end{abstract}

@@

\begin{figure}
    \includegraphics[width=.6\textwidth]{fig-topology.pdf}
    \caption{Decay topology for decays $B\to P_1 P_2 \ell_1 \bar\ell_2$. \label{fig:topology}}
\end{figure}

@@

\subsection{Recipe for Including Detector Effects}
\label{sec:systematics:acceptance}

@@

The accuracy of the unfolding process as outlined above critically depends on both the accuracy
of the detector simulation, as well as the uncertainties induced by the MC estimates.
For an experimental analysis, one would now turn to the determination of the distribution of $\vecest{S}$
as a function of the detector setup. This would involve the determination of both $M$ and $\vecest{Q}$ for a number of
detector configurations, and subsequent profiling or marginalization.
While such considerations of any detector simulation are beyond the scope of this work, we can, however,
comment on the MC-induced uncertainties. As usual, one needs to find a balance between
compute time and accuracy. For a uniangular distribution and a total of $6\cdot 10^6$ MC
samples (i.e., $2 \cdot 10^6$ MC samples per row of $M_{ij}$), we find that the error on the mean is
smaller than $6\cdot 10^{-4}$. Let us conclude this section by commenting that parts of the unfolding
matrix are universal in a sense. Specifically, they can be reused in analyses with a similar underlying decay.
Therefore, computing resources spent on improving its accuracy are not wasted.

@@

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S5.png}
            \includegraphics[width=0.45\textwidth]{figs/error-Q2_5_6_S7.png}
            \caption{Uncertainty of the angular observables $S_5$ and $S_7$, extracted from simulated events of the decay $B\to K^*\ell^+\ell^-$ and as a function of the number of simulated events $K$. The red curve represents a fit to the function given in \refeq{unc-on-mean}.}
            \label{fig:errors}
\end{figure}

@@

From the positivity of amplitudes and unitarity, one immediately finds
\begin{equation}
    \int \rmdx{\cos\theta}  \, |A_i(\theta)|^2 = \sum_{l_i = 0}^\infty |A_i^{(l_i)}|^2 n_{l_i, m_i} < \infty\,,
\end{equation}
where we introduce $n_{l,m}$ via the scalar product of two associated Legendre polynomials,
\begin{equation}
    \label{eq:legendre-scalar-product}
    n_{l, m} \delta_{l, l'} \equiv \int_{-1}^1 \rmdx{\cos\theta} p_{l}^{(m)}(\cos\theta) p_{l'}^{(m)} (\cos\theta)  = \frac{2}{(2 l + 1)} \frac{(l + m)!}{(l - m)!} \delta_{l, l'}\,.
\end{equation}
Using the large-$l$ behaviour of $n_{l,m}$ one finds that with
increasing $l_i$ the partial-wave amplitudes must asymptotically fall
off as
\begin{equation}
    |A_i^{(l_i)}| \Big|_{l_i \gg m_i} \sim \frac{1}{\sqrt{l_i}}\,.\todo{wrong!}
\end{equation}
This motivates to cut off the partial-wave expansion at some arbitrary angular momentum $L$.\\

@@

In the last step, we use Gaunt's formula \cite{Gaunt:1929} to integrate a triple product of associated Legendre polynomials,
\begin{equation}
\begin{aligned}
    T_{l_1,l_2,j}^{(m_1,m_2)}
        & = \int_{-1}^{+1} \rmdx{\cos\theta} p_{j}^{(m_1 + m_2)}(\cos\theta) p_{l_1}^{(m_1)}(\cos\theta) p_{l_2}^{(m_2)}(\cos\theta)\\
        & = (-1)^{s - l_1 - m_2} \frac{2 (l_1 + m_1)! (l_2 + m_2)! (2s - 2 l_2)! s!}{(l_1 - m_1)! (s - j)! (s - l_1)! (s - l_2)! (2s + 1)!}\\
        & \times \sum_{t=p}^q (-1)^t \frac{(j + m + t)!(l_1 + l_2 - m - t)!}{t! (j - m - t)! (l_1 - l_2 + m + t)! (l_2 - m_2 - t)!}\,,
\end{aligned}
\end{equation}
where
\begin{equation}
\begin{aligned}
    m & = m_1 + m_2\,, &
    m_1 & \geq m_2\,,  \\
    j, l_1, l_2 & \geq 0\,, &
    m, m_1, m_2 & \geq 0\,,
\end{aligned}
\end{equation}
and
\begin{equation}
\begin{aligned}
    s & = \frac{j + l_1 + l_2}{2}\,, &
    p & = \max(0, l_2 - l_1 - m)\,, &
    q & = \min(l_1 + l_2 - m, j - m, l_2 - m_2)\,.
\end{aligned}
\end{equation}
The necessary conditions for $T \neq 0$ are
\begin{equation}
    \label{eq:angular-momentum-addition}
    s \in \mathbb{N}\qquad \wedge \qquad l_1 - l_2 \leq l \leq l_1 + l_2,.
\end{equation}
The latter condition is well known from the addition rules of angular momenta. Note, however, that
the sum in \refeq{def-partial-wave-observable} goes to infinitely high angular momenta $l_1$ and $l_2$. As a consequence
of this and of \refeq{angular-momentum-addition}, the angular observables $P_{j,m}$
consist of sums with infinitely many terms. It is then up to theoretical analyses to
estimate or calculate the impact of individual contributions, with or without the cutoff $L$.
We close by remarking that for any angular-momentum cutoff $l < L < \infty$, one can find an upper limit for the squared sum of all remaining partial waves through a comparison with the inclusive decay
width.
